TRIK Studio: Technical Introduction

### 2017-02-09 2017-trikStudioEnglish/trikStudioEnglish.tex
translated generators

The the first mode the program is being interpreted on a two-dimensional robot model (see Section~\ref{chapter:2dModel}). Users are able to create 2D-model of the world surrounding the robot out of walls, colored elements and floor markup. According to TRIK Studio users this feature is very useful for initial program debugging, before any interaction with real robots. Our experience shows that this virtual model editor allows to recreate most of fields and obstacle courses used in competitions in robotics. Having such a simulation environment makes possible learning robotics and programming even without having the real robotic kits. There is also an experimental support for V-Rep 3D visual simulation environment~\cite{rohmer2013v}.

@@

This section discusses the subsystem handling TRIK Studio interaction with robots over different protocols. It is one of the lowest abstraction layers shown on the Fig.~\ref{image:commonTSArch}. Its API provides the following operations:

@@

Одна из наиболее важных и востребованных функций среды TRIK Studio --- генерация читаемого кода по визуальной диаграмме. По одной и той же диаграмме может быть сгенерирован код на любом поддерживаемом средой текстовом языке (C, JavaScript, F\#, Pascal, РуСи~\cite{тереховотечественные}, байткод ВМ EV3). Под визуальными диаграммами в данном разделе будем подразумевать программы на языке для начинающих (с передачей управления), под генераторами --- модули генерации кода по визуальным диаграммам именно этого языка.

Визуальная диаграмма состоит из блоков и стрелок, таким образом, описывает \textit{граф потока управления} программы~\cite{steven1997advanced}. Все основные алгоритмические конструкции (развилки, циклы, конструкции выбора и т.д.) складываются из стрелок. Например, чтобы задать бесконечный цикл, достаточно лишь провести стрелку от блока к одному из предыдущих, для описания циклов вида \texttt{while-do}, \texttt{do-while} или с инструкцией \texttt{break} внутри тела достаточно лишь провести одну из веток развилки из тела цикла к соответствующему блоку вне его. Очевидно, что одна стрелка на диаграмме соответствует инструкции \texttt{goto} в коде. Однако код, содержащий инструкции \texttt{goto} весьма труден для чтения человеком, тем более не подходит для обучения азам текстового программирования, поэтому следует избегать их появления в сгенерированном коде. Тем не менее, на TRIK Studio возможно написать программу, не выразимую общепринятыми алгоритмическими конструкциями, в таком случае появления инструкций \texttt{goto} в коде не избежать (существуют методы автоматического преобразования программы с goto в структурную программу, но результат их работы не лучше с точки зрения обучения текстовому программированию). С другой стороны, далеко не все современные текстовые языки поддерживают инструкцию \texttt{goto} (например, в языке JavaScript такой поддержки нет). Будем называть \textit{приводимыми} диаграммы, которые можно выразить стандартными алгоритмическими блоками (\texttt{if-then}, \texttt{if-then-else}, \texttt{while-do loop}, \texttt{do-while loop}, \texttt{while-break loop}, \texttt{switch}), код на текстовом языке, соответствующий структурированной диаграмме также будем называть \textit{структурированным}. В противном случае будем говорить, что диаграмма \textit{неприводима}, а код будем называть \textit{неструктурированным}. В случае, если диаграмму не удается сгенерировать в структурированный код, пользователю должно быть показано предупреждение.

Таким образом, реализация системы генераторов в TRIK Studio потребовала решения двух нетривиальных задач, которые сформулированы ниже в виде требований.

\begin{itemize}
    \item Система генераторов должна быть организована таким образом, чтобы добавление в систему генератора в новый текстовый язык занимало минимальное количество времени и усилий. По возможности, это должно быть по силам даже людям, не знакомым с кодом системы.
    \item Для каждого языка (где это возможно) система должна поддерживать два режима генерации: приводимые диаграммы должны быть сгенерированы в структурированный код, неприводимые --- в неструктурированный. При этом успешность структурирования диаграммы не должна зависеть от ее размера и сложности.
\end{itemize}

Первая задача является чисто архитектурной. Ее решение представлено на рисунке~\ref{image:generatorsArchitecture}. Основная идея --- разделение процесса генерации на два этапа. На первом диаграмма преобразуется в представление, независимое от целевого языка генерации --- \textit{семантическое дерево}. Семантическое дерево упорядочивает структуру графа потока управления до <<древесной>>, в независимости от того, является ли диаграмма приводимой или нет. В случае, если диаграмма приводима, родительский узел семантического дерева всегда соответствует блоку вычислений в целевом коде, дети --- инструкциям этого блока, среди которых нет \texttt{goto}. Опишем последовательность действий, отображенных на рисунке~\ref{image:generatorsArchitecture}.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{TS_Generators_Architecture.png}
    \caption{Архитектура подсистемы генерации кода в TRIK Studio}
    \label{image:generatorsArchitecture}
\end{figure*}

На вход генератору поступает визуальная диаграмма. На первом этапе диаграмма обходится в глубину компонентой-валидатором (построенной с применением шаблона проектирования <<посетитель>>). Для всех выражений на встроенном текстовом языке в свойствах блоков происходит их синтаксический разбор и вывод типов. В случае, если диаграмма или код в свойстве содержит ошибки, о них сообщается пользователю в локализованном виде и процесс завершается. В случае, если диаграмма синтаксически корректна, она поступает на вход анализатору потока управления, о котором будет рассказано ниже. В зависимости от того, может ли диаграмма быть сгенерирована в структурированный код, выбирается промежуточный генератор независимого представления, который извлекает семантическое дерева из потока управления диаграммы. Если целевой язык не поддерживает инструкцию \texttt{goto} или, наоборот, высокоуровневые конструкции, как, например, ассемблер-подобный байткод ВМ EV3 (что указывается в конфигурации генератора), соответствующему генератору будет предпочтена его альтернатива. Наконец, на последнем шаге процесса генерации семантическое дерево печатается в целевой текстовый язык. Для этого используется подход с заданием шаблонов генерации для целевого текстового языка. Все, что необходимо сделать для реализации нового генератора --- переопределить набор этих шаблонов и указать конфигурацию генератора. К примеру, шаблон генерации условного оператора в язык C выглядит следующим образом: \begin{verbatim}
    if (@@CONDITION@@) {
        @@THEN_BODY@@
    } else {
        @@ELSE_BODY@@
    }
\end{verbatim}

Вторая из решенных проблем --- модуль анализатора потока управления. В терминах текстовых языков задача звучит так: по данной программе, поток управления которой описан исключительно инструкциями \texttt{goto}, требуется выдать эквивалентную структурированную программу. Данная задача решалась различными исследователями, работающими в области декомпиляции программного кода~\cite{steven1997advanced,деревенец2009структурный}. Приемлемый способ ее решения по методу \textit{интервального анализа} графа потока управления описан, к примеру, в монографии~\cite{steven1997advanced}. Алгоритм интервального анализа графа потока управления диаграммы с незначительными модификациями реализован в TRIK Studio. Коротко опишем его.

Неформально говоря, \textit{интервал} --- это участок графа потока управления с одним входом и одним выходом. Целью алгоритма является построение дерева вложенности интервалов графа потока управления программы, которое в нашем случае и является семантическим деревом. Алгоритм работает на рекурсивном подъеме поиска в глубину в графе потока управления. На каждом шаге алгоритм пробует сопоставить подграф, выходящий из текущей вершины, с каждым из шаблонов элементарных интервалов (алгоритмических конструкций, в терминах которых и будет структурирована программа). Если такой шаблон удается найти, весь подграф сворачивается в одну вершину, и процесс продолжается. Самый простой пример такого шаблона --- два интервала, соединенные ровно одной дугой, что соответствует последовательной композиции. Если на каком-то шаге ни один шаблон не подошел, то диаграмма считается неприводимой.

Очевидным ограничениям на целевой язык в текущей реализации генераторов является его императивность: цепочка блоков на диаграмме должна преобразовываться в последовательное исполнение утверждений в текстовом языке, что проще всего достигается применением оператора последовательной композиции. Тем не менее, идея использования промежуточного шага генерации в независимое представление работает и здесь. К примеру, для поддержки генерации в чисто функциональные языки достаточно расширить элементы независимого представления конструкцией продолжения (continuation passing style, CPS) и добавить промежуточный генератор диаграммы в независимое представление в форме CPS. После этого достаточно переопределить шаблоны печати независимого представления во все необходимые языки.

>>>>

The the first mode the program is being interpreted on a two-dimensional robot model (see Section~\ref{chapter:2dModel}). Users are able to create 2D-model of the world surrounding the robot from walls, colored elements and floor markup. According to TRIK Studio users this feature is very useful for initial program debugging, before any interaction with real robots. Our experience shows that this virtual model editor allows to recreate most of fields and obstacle courses used in competitions in robotics. Having such a simulation environment makes possible learning robotics and programming even without having the real robotic kits. There is also an experimental support for V-Rep 3D visual simulation environment~\cite{rohmer2013v}.

@@

This section discusses the subsystem handling TRIK Studio interaction with robots over different protocols. It is one of the lowest abstraction layers shown in Fig.~\ref{image:commonTSArch}. Its API provides the following operations:

@@

One of the most interesting and demanded TRIK Studio features is generation of well-readable textual code from created diagrams. At the moment of writing several languages are supported for code generation: C, JavaScript, F\#, Pascal, Russian version of C~\cite{тереховотечественные} and byte code for Lego EV3 virtual machine. In this section we describe only code generators for control-flow programs.

A visual diagram is built from blocks and arrows, defining program's \textit{control flow graph}~\cite{steven1997advanced}. All algorithmic operators like conditions, loops, switches, etc. are defined by arrows. For instance, to create an infinite loop one should create an arrow from a block to one of the previous ones, to create \texttt{while-do} and \texttt{do-while} loops or loops with \texttt{break} statement inside it is enough to simply connect one block inside a loop with a block from the outside. It is obvious that each arrow on the diagram corresponds to a \texttt{goto} statement in the code. But the code containing \texttt{goto} statements is very hard to understand, and especially in not suitable for teaching the basics of programming, so we decided to avoid using them in generated code as much as possible. We cannot eliminate \texttt{goto} statements at all because it is possible to create a TRIK Studio diagram that could not be expressed using standard algorithmic operators (there are algorithms for automatic translation of \texttt{goto}-containing programs into structural programs, but the result is not much better from the teaching programming point of view). On the other hand, not all modern textual programming languages support \texttt{goto} statements (for example, JavaScript does not). We call \textit{structural} diagrams that could be expressed using structural algorithmic statements (\texttt{if-then}, \texttt{if-then-else}, \texttt{while-do loop}, \texttt{do-while loop}, \texttt{while-break loop}, \texttt{switch}). Textual code generated from them we also call \textit{structural}. Otherwise we say that the diagram (and the code generated from it) is \textit{not structural}. If structural code could not be generated from a given diagram, users are notified with a warning.

Considering these issues implementation of code generation subsystem in TRIK Studio required solving two non-trivial tasks, given below in a form of requirements.

\begin{itemize}
    \item Code generation subsystem should be organized in such a way to minimize time and effort for adding of a new generator. Preferably is should be possible even for an external developer to do it.
    \item For each language (where applicable) the subsystem should support two generation modes: structural diagrams should be translated into structural code, not structural diagrams -- into not structural code. The success of structuring the diagram should not depend on its size and complexity.
\end{itemize}

The first task is purely architectural one. The way to solve it is shown on Fig.~\ref{image:generatorsArchitecture}. The main idea here is to divide code generation process into two steps. First, the diagram is translated into a representation independent of a target language --- a \textit{semantic tree}. Semantic trees reorder graph model of the control flow into a tree model regardless of whether the diagram is structural or not. If it is, a parent node in a semantic tree always corresponds to an execution block in generated code, and its children nodes correspond to statements of this block, which exclude \texttt{goto}. Let's outline the code generation process shown in Fig.~\ref{image:generatorsArchitecture}.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{TS_Generators_Architecture.png}
    \caption{Code generation subsystem's architecture}
    \label{image:generatorsArchitecture}
\end{figure*}

A code generator gets a diagram as input data. First, the diagram is traversed in depth-first order by a validation component (build based on the <<Visitor>> design pattern). Then, for each computable expression used inside block property values parsing and type inference procedure is performed. If the diagram or any code in any property value contains an error, users are notified and the code generation process terminates. It the diagram is syntactically correct, it is passed to a control flow analyzer, that is described below. Depending on if the diagram can be translated into structural code or not, an intermediate generator of code-independent diagram representation is selected, which extracts the semantic tree from the control flow graph. Finally, the semantic tree is printed into a target textual language. To achieve it, a template-based approach is used, so to create a new generator a programmer have to provide a set of such code templates and define generator specification. For example, generation templates for condition expression if C looks like this:
\begin{verbatim}
    if (@@CONDITION@@) {
        @@THEN_BODY@@
    } else {
        @@ELSE_BODY@@
    }
\end{verbatim}

The second task is solved by a control flow analyzer module. In terms of textual language the task is rephrased like this: by a given program containing \texttt{goto} statements create an equivalent structured program. This problem was solved by researchers in a field of code decompilation~\cite{steven1997advanced,деревенец2009структурный}. For example, such an algorithm based on interval analysis of control flow graph is presented in~\cite{steven1997advanced}. With minor modifications if was implemented in TRIK Studio. Let's describe it briefly.

Informally speaking, an \textit{interval} is a part of a control flow graph with one input and one output. The algorithm aims to build a nesting tree for all graph intervals and is based on recursive ascend of depth-first search. On each step the algorithm tries to match a subgraph, outgoing from the current node, with each of basic interval templates (algorithmic statements that are used to structure the program). If such a template is found, the whole subgraph is folded into a single node and the process continues. The most simple example of such a template --- two intervals connected with an only one edge, which matches with sequential composition. If on any step no template could be found, the diagram is considered not structural.

One obvious limitation of the target textual language for code generation is that it should be an imperative one: the chain of blocks must be translated into a sequence of target language expressions, which is better achieved using sequential composition. Nevertheless, an approach with intermediate step with generation into intermediate representation could work here too. For example, to support code generation into pure functional language we could add a continuation element into intermediate representation and add intermediate generator from visual diagrams into continuation passing style (CPS) representation. After that code generation templates should be updated to support all necessary languages.
